The Leaky Semicolon (Supplementary Material)

### 2021-11-10 paper/mca1.tex
removing todos. Integrated consistency and causal strengtheining.
accepted!
pass through 4-5

In this section, we develop a model of concurrent computation by adding
% \emph{parallel composition} and 
\emph{reads-from} to \reffig{fig:seq}.  To model coherence and
synchronization, we add \emph{delay} to the rule for sequential composition.
For \mca{} architectures, it is sufficient to encode delay in the pomset
order.  The resulting model, \PwTmca{1}, supports optimal lowering for
relaxed access on \armeight{}, but requires extra synchronization for
acquiring reads.\footnote{\citet{DBLP:journals/pacmpl/JagadeesanJR20}
  erroneously elide the required synchronization on acquiring reads.}
(\emph{Lowering} is the translation of language-level operators to machine
instructions.  A lowering is \emph{optimal} if it provides the most efficient
execution possible.)

A variant, \PwTmca{2}, supports optimal lowering for all access modes on
\armeight{}.  To achieve this, \PwTmca{2} drops the global requirement that
\emph{reads-from} implies pomset order \eqref{pom-rf-le}.  The models are the
same, except for \emph{internal reads}, where a thread reads its own write.
We show an example at the beginning of \textsection\ref{sec:mca2}.

The lowering proofs can be found in \cite{appendix}.  The proofs
use recent alternative characterizations of \armeight{} \cite{armed}.

@@

  \noindent
  \label{def:pwt:mca:complete}
  The definition of completeness extends \refdef{def:pomset} as follows:
  \begin{multicols}{2}
    \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
      \setcounter{enumi}{\value{Brf}}
    \item \label{top-rf}
      if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
    \end{enumerate}
  \end{multicols}

  \noindent
  The semantic function extends \reffig{fig:seq} as follows:
  \begin{multicols}{2}
    \begin{enumerate}[topsep=0pt,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
    \item[{\labeltext[\textsc{p}7]{(\textsc{p}7)}{par-rf}}]
      \labeltext[\textsc{s}7]{(\textsc{s}7)}{seq-rf}\;
      \labeltext[\textsc{i}7]{(\textsc{i}7)}{if-rf}\;
      ${\rrfx}\rextends{\rrfx_1}{\rrfx_2}$,
      \setcounter{enumi}{\value{le}}
    \item[]
      \begin{enumerate}[leftmargin=0pt]
      \item \label{seq-le-delays}
        if $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ then $\bEv\le\aEv$.
      \end{enumerate}
    \end{enumerate}  
  \end{multicols}
  % The semantic rules are as follows:
  % \input{fig-mca1.tex}
\end{definition}
% Let
% \begin{math}
%   \sem{\aCmd_1 \RPAR \aCmd_2} = \sPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
% \end{math}
We write $\semmca{1}{}$ for the semantic function when it is unclear from context.

In complete pomsets, $\rrfx$ must pair every read with a matching write
\eqref{top-rf}.  The requirements \ref{pom-rf-match}, \ref{pom-rf-block}, and
\ref{pom-rf-le} guarantee that reads are \emph{fulfilled}, as in
\cite[\textsection 2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.

The semantic rules are mostly straightforward: Parallel composition is disjoint
union, and all constructs respect reads-from.  The monoid laws
(\reflem{lem:monoid}) extend to parallel composition, with $\SKIP$ as right
unit only due to the asymmetry of \ref{par-tau}.

Only \ref{seq-le-delays} requires explanation.
% \begin{lemma}
%   \label{lem:rf:implies:le}
%   For any $\aPS$ in the range of $\sembase{}$, $\bEv\xrfx\aEv$ implies
%   $\bEv\le\aEv$.
%   \vspace{-.5\baselineskip}
%   \begin{proof}
%     Induction on the definition of $\sembase{}$, using \ref{par-rf-le1}, \ref{par-rf-le2}, and \ref{pom-rf-le}.
%   \end{proof}
% \end{lemma}
% \footnote{The
%   basic model would be the same if we move $\rrfx$ from the model itself to
%   be existentially quantified in the definition of top-level pomset, along
%   with \ref{pom-rf-match} and \ref{pom-rf-block}.  This was the approach of
%   \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} We include $\rrfx$
%   explicitly for use in \textsection\ref{sec:arm2}, where we introduce a
%   variant semantics $\frf{\semrr{}}$ where \ref{pom-rf-le} is not required.}
From \refdef{def:actions}, recall that $\aAct \rdelaysdef \bAct$ if
$\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or
$\aAct\eqreordersc\bAct$.  \ref{seq-le-delays} guarantees that sequential
order is enforced between conflicting accesses of the same location
($\eqreorderco$), into a release and out of an acquire ($\reorderra$), and
between SC accesses ($\eqreordersc$).  Combined with the fulfillment
requirements (\ref{pom-rf-match}, \ref{pom-rf-block} and \ref{pom-rf-le}),
these ensure coherence, publication, subscription and other idioms.  For
example, consider the following:\footnote{We use different colors for arrows representing order:
  \begin{multicols}{2}
    \begin{itemize}  
    \item \makebox{$\bEv\xwki\aEv$} arises from $\eqreorderco$ \eqref{seq-le-delays},
    \item \makebox{$\bEv\xsync\aEv$} arises from $\reorderra$ or $\eqreordersc$ \eqref{seq-le-delays},
    \item \makebox{$\bEv\xpo\aEv$} arises from \makebox[0pt][l]{control/data/address \emph{dependency} (\ref{seq-kappa}, definition of $\labelingForm'_2(\bEv)$),}
    \item \makebox{$\bEv\xrf\aEv$} arises from \emph{reads-from} \eqref{pom-rf-match},
      % \item \makebox{$\bEv\xsyncsc\aEv$} arises from \emph{strong fencing} \eqref{cand-lesync-sc},
    \item \makebox{$\bEv\xwk\aEv$} arises from \emph{blocking} \eqref{pom-rf-block}.
    \end{itemize}    
  \end{multicols}
  In \PwTmca{2}, it is possible for $\rrf$ to contradict
  $\lt$.  In this case, we use a dotted arrow for $\rrf$: $\bEv\xrfint\aEv$
  indicates that $\aEv\lt\bEv$.}
\begin{gather*}
  \taglabel{pub}
  \begin{gathered}    
    \PW{x}{0}\SEMI 
    \PW{x}{1}\SEMI \PW[\mREL]{y}{1} \PAR \PR[\mACQ]{y}{r}\SEMI \PR{x}{s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy1}{\DW[\mREL]{y}{1}}{right=of wx1}
        \event{ry1}{\DR[\mACQ]{y}{1}}{right=2.5em of wy1}
        \event{rx0}{\DR{x}{0}}{right=of ry1}
        \sync{wx1}{wy1}
        \sync{ry1}{rx0}
        \rf{wy1}{ry1}
        \wk[out=-170,in=-10]{rx0}{wx1}
        \wki{wx0}{wx1}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
The execution is disallowed due to the cycle.  All of the order shown is
required at top-level: The intra-thread order comes from \ref{seq-le-delays}:
$\DWP{x}{0}\xwki \DWP{x}{1}$ is required by $\eqreorderco$.
$\DWP{x}{1}\xsync \DWP[\mREL]{y}{1}$ and $\DRP[\mACQ]{y}{1}\xsync \DRP{x}{0}$ are
required by $\reorderra$.  The cross-thread order is required by fulfillment:
\ref{top-rf} requires that all top-level reads are in the image of $\xrfx$.
\ref{pom-rf-match} ensures that $\DWP[\mREL]{y}{1}\xrfx \DRP[\mACQ]{y}{1}$, and
\ref{pom-rf-le} subsequently ensures that
$\DWP[\mREL]{y}{1}\lt \DRP[\mACQ]{y}{1}$.  The \emph{antidependency}
$\DRP{x}{0}\xwk \DWP{x}{1}$ is required by \ref{pom-rf-block}.  (Alternatively,
we could have $\DWP{x}{1}\xwk \DWP{x}{0}$, again resulting in a cycle.)

>>>>

In this section, we develop a model of concurrent computation by adding
% \emph{parallel composition} and 
\emph{reads-from} to \reffig{fig:seq}.  To model coherence and
synchronization, we add \emph{delay} to the rule for sequential composition.
For \mca{} architectures, it is sufficient to encode delay in the pomset
order.  The resulting model, \PwTmca{1}, supports optimal lowering for
relaxed access on \armeight{}, but requires extra synchronization for
acquiring reads.  (\emph{Lowering} is the translation of language-level
operators to machine instructions.  A lowering is \emph{optimal} if it
provides the most efficient execution possible.)

A variant, \PwTmca{2}, supports optimal lowering for all access modes on
\armeight{}.  To achieve this, \PwTmca{2} drops the global requirement that
\emph{reads-from} implies pomset order \eqref{pom-rf-le}.  The models are the
same, except for \emph{internal reads}, where a thread reads its own write.
We show an example at the beginning of \textsection\ref{sec:mca2}.
%
The lowering proofs can be found in \refAppendix{sec:arm}.  The proofs
use recent alternative characterizations of \armeight{} \cite{armed}.

@@

  \label{def:pwt:mca:complete}
  \noindent
  \begin{minipage}{1.0\linewidth}
    The definition of completeness extends \refdef{def:pomset} as follows:
    \begin{multicols}{2}
      \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
        \setcounter{enumi}{\value{Brf}}
      \item \label{top-rf}
        if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
      \end{enumerate}
    \end{multicols}
  \end{minipage}
  \smallskip

  \noindent
  \begin{minipage}{1.0\linewidth}
    The semantic function extends \reffig{fig:seq} as follows:
    \begin{multicols}{2}
      \begin{enumerate}[topsep=0pt,label=(\textsc{i}\arabic*),ref=\textsc{s}\arabic*]
        \setcounter{enumi}{\value{le}}
      \item[]
        \begin{enumerate}[leftmargin=0pt]
        \item \label{seq-le-delays}
          if $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ then $\bEv\le\aEv$,
        \end{enumerate}
      \item[{\labeltext[\textsc{p}7]{(\textsc{p}7)}{par-rf}}]
        \labeltext[\textsc{s}7]{(\textsc{s}7)}{seq-rf}\;
        \labeltext[\textsc{i}7]{(\textsc{i}7)}{if-rf}\;
        ${\rrfx}\rextends{\rrfx_1}{\rrfx_2}$.
      \end{enumerate}  
    \end{multicols}
  \end{minipage}
  \smallskip
  % The semantic rules are as follows:
  % \input{fig-mca1.tex}
\end{definition}
% Let
% \begin{math}
%   \sem{\aCmd_1 \RPAR \aCmd_2} = \sPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
% \end{math}
%We write $\semmca{1}{}$ for the semantic function when it is unclear from context.

In complete pomsets, reads-from ($\rrfx$) must pair every read with a
matching write \eqref{top-rf}.  The requirements \ref{pom-rf-match},
\ref{pom-rf-block}, and \ref{pom-rf-le} guarantee that reads are
\emph{fulfilled}, as in \cite[\textsection
2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.  Parallel composition, sequential
composition, and the conditional respect reads-from (\ref{par-rf},
\ref{seq-rf}, \ref{if-rf}).

% The semantic rules are mostly straightforward: Parallel composition is disjoint
% union, and all constructs respect reads-from.  The monoid laws
% (\reflem{lem:monoid}) extend to parallel composition, with $\SKIP$ as right
% unit only due to the asymmetry of \ref{par-tau}.

%Only \ref{seq-le-delays} requires explanation.
% \begin{lemma}
%   \label{lem:rf:implies:le}
%   For any $\aPS$ in the range of $\sembase{}$, $\bEv\xrfx\aEv$ implies
%   $\bEv\le\aEv$.
%   \vspace{-.5\baselineskip}
%   \begin{proof}
%     Induction on the definition of $\sembase{}$, using \ref{par-rf-le1}, \ref{par-rf-le2}, and \ref{pom-rf-le}.
%   \end{proof}
% \end{lemma}
% \footnote{The
%   basic model would be the same if we move $\rrfx$ from the model itself to
%   be existentially quantified in the definition of top-level pomset, along
%   with \ref{pom-rf-match} and \ref{pom-rf-block}.  This was the approach of
%   \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} We include $\rrfx$
%   explicitly for use in \textsection\ref{sec:arm2}, where we introduce a
%   variant semantics $\frf{\semrr{}}$ where \ref{pom-rf-le} is not required.}
From \refdef{def:actions}, recall that $\aAct \rdelaysdef \bAct$ if
$\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or
$\aAct\eqreordersc\bAct$.  \ref{seq-le-delays} guarantees that sequential
order is enforced between conflicting accesses of the same location
($\eqreorderco$), into a release and out of an acquire ($\reorderra$), and
between SC accesses ($\eqreordersc$).  Combined with the fulfillment
requirements (\ref{pom-rf-match}, \ref{pom-rf-block}, \ref{pom-rf-le}),
these ensure coherence, publication, subscription and other idioms.  For
example, consider the following:\footnote{We use different colors for arrows representing order:
  \begin{multicols}{2}
    \begin{itemize}  
    \item \makebox{$\bEv\xwki\aEv$} arises from $\eqreorderco$ \eqref{seq-le-delays},
    \item \makebox{$\bEv\xsync\aEv$} arises from $\reorderra$ or $\eqreordersc$ \eqref{seq-le-delays},
    \item \makebox{$\bEv\xpo\aEv$} arises from \makebox[0pt][l]{control/data/address \emph{dependency} (\ref{seq-kappa}, definition of $\labelingForm'_2(\bEv)$),}
    \item \makebox{$\bEv\xrf\aEv$} arises from \emph{reads-from} \eqref{pom-rf-match},
      % \item \makebox{$\bEv\xsyncsc\aEv$} arises from \emph{strong fencing} \eqref{cand-lesync-sc},
    \item \makebox{$\bEv\xwk\aEv$} arises from \emph{blocking} \eqref{pom-rf-block}.
    \end{itemize}    
  \end{multicols}
  In \PwTmca{2}, it is possible for $\rrf$ to contradict
  $\lt$.  In this case, we use a dotted arrow for $\rrf$: $\bEv\xrfint\aEv$
  indicates that $\aEv\lt\bEv$.}
\begin{gather*}
  \taglabel{pub}
  \begin{gathered}    
    \PW{x}{0}\SEMI 
    \PW{x}{1}\SEMI \PW[\mREL]{y}{1} \PAR \PR[\mACQ]{y}{r}\SEMI \PR{x}{s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy1}{\DW[\mREL]{y}{1}}{right=of wx1}
        \event{ry1}{\DR[\mACQ]{y}{1}}{right=2.5em of wy1}
        \event{rx0}{\DR{x}{0}}{right=of ry1}
        \sync{wx1}{wy1}
        \sync{ry1}{rx0}
        \rf{wy1}{ry1}
        \wk[out=-170,in=-10]{rx0}{wx1}
        \wki{wx0}{wx1}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
The execution is disallowed due to the cycle.  All of the order shown is
required at top-level: The intra-thread order comes from \ref{seq-le-delays}:
$\DWP{x}{0}\xwki \DWP{x}{1}$ is required by $\eqreorderco$.
$\DWP{x}{1}\xsync \DWP[\mREL]{y}{1}$ and $\DRP[\mACQ]{y}{1}\xsync \DRP{x}{0}$ are
required by $\reorderra$.  The cross-thread order is required by fulfillment:
\ref{top-rf} requires that all top-level reads are in the image of $\xrfx$.
\ref{pom-rf-match} ensures that $\DWP[\mREL]{y}{1}\xrfx \DRP[\mACQ]{y}{1}$, and
\ref{pom-rf-le} subsequently ensures that
$\DWP[\mREL]{y}{1}\lt \DRP[\mACQ]{y}{1}$.  The \emph{antidependency}
$\DRP{x}{0}\xwk \DWP{x}{1}$ is required by \ref{pom-rf-block}.  (Alternatively,
we could have $\DWP{x}{1}\xwk \DWP{x}{0}$, again resulting in a cycle.)

### 2021-11-10 paper/mca2.tex
pass through 4-5

Lowering \PwTmca{1} to \armeight{} requires a full fence after every
acquiring read.  To see why, consider the following attempted
execution, where the final values of both $x$ and $y$ are $2$.
\begin{gather*}
  \taglabel{internal-acq}
  \begin{gathered}
  \PW{x}{2}\SEMI 
  \PR[\mACQ]{x}{r}\SEMI
  \PW{y}{r{-}1} \PAR
  \PW{y}{2}\SEMI
  \PW[\mREL]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{2}}{}
      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \event{d}{\DW{y}{2}}{right=2.5em of c}
      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
      \rfint{a}{b}
      \sync{b}{c}
      \wk{c}{d}
      \sync{d}{e}
      \wk[out=-170,in=-10]{e}{a}
    \end{tikzinline}}
\end{gathered}
\end{gather*}
The execution is  allowed by \armeight, but disallowed by \PwTmca{1}, due to
the cycle.

>>>>

Lowering \PwTmca{1} to \armeight{} requires a full fence before every
acquiring read.\footnote{\citet{DBLP:journals/pacmpl/JagadeesanJR20}
  erroneously elide the required synchronization on acquiring reads.}  To see
why, consider the following attempted execution, where the final values of
both $x$ and $y$ are $2$.
\begin{gather*}
  \taglabel{internal-acq}
  \begin{gathered}
  \PW{x}{2}\SEMI 
  \PR[\mACQ]{x}{r}\SEMI
  \PW{y}{r{-}1} \PAR
  \PW{y}{2}\SEMI
  \PW[\mREL]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{2}}{}
      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \event{d}{\DW{y}{2}}{right=2.5em of c}
      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
      \rfint{a}{b}
      \sync{b}{c}
      \wk{c}{d}
      \sync{d}{e}
      \wk[out=-170,in=-10]{e}{a}
    \end{tikzinline}}
\end{gathered}
\end{gather*}
The execution is  allowed by \armeight, but disallowed by \PwTmca{1}, due to
the cycle.

### 2021-11-10 paper/outro.tex
pass through 4-5

We have defined the soundness of compiler optimizations in the model, rather
than contextually: $\aCmd'$ is a sound refinement of $\aCmd$ if
$\sem{\aCmd'}\subseteq\sem{\aCmd}$.  This approach has several
advantages---for example, it is immediate that a sound optimization is sound
in any context.  It also has a disadvantage: some optimizations complicate
the semantics.  For example, \PwTmca{} does not validate access elimination,
such as store-forwarding and dead-write-removal---consider that complete
executions of
\begin{math}
  \sem{\PW{x}{1}\SEMI\PR{x}{r}}
\end{math}
must include a read action and that complete executions of
\begin{math}
  \sem{\PW{x}{1}\SEMI\PW{x}{2}}
\end{math}
must include two write actions.  As another example, \PwTmca{} does not
validate the reverse inclusions for
\reflem{lem:if}\eqref{lem:ifelse:if:if}---consider that
\begin{math}
  \sem{\IF{r}
  \THEN \PW{x}{1}
  \ELSE \PW{x}{2}
  \FI}
\end{math}
has an augmented (\reflem{lem:augment}) execution with
\begin{math}
  (r{=}0\mid\DW{x}{2})
  \xwki
  (r{\neq}0\mid\DW{x}{1}),
\end{math}
whereas
\begin{math}
  \sem{
    \IF{r} \THEN \PW{x}{1}\FI
    \SEMI
    \IF{\BANG r} \THEN \PW{x}{2}\FI
  \FI}
\end{math}
has no such execution.  We expect that these optimizations can be validated,
at the cost of complicating the semantics.  For access elimination, it is
likely sufficient to allow events with different actions to merge.  For
\reflem{lem:if}\eqref{lem:ifelse:if:if}, it is likely sufficient to encoding
$\rdelay$ in the logic---the problem in the execution above is that $\rdelay$
introduces order regardless of whether preconditions are disjoint.
%\todo{Comment on \PwTpo{} and \PwTc{}?}

>>>>

We have defined the soundness of compiler optimizations in the model, rather
than contextually: $\aCmd'$ is a sound refinement of $\aCmd$ if
$\sem{\aCmd'}\subseteq\sem{\aCmd}$.  This approach has several
advantages---for example, it is immediate that a sound optimization is sound
in any context.  It also has a disadvantage: some optimizations complicate
the semantics.  For example, \PwTmca{} does not validate access elimination,
such as store-forwarding and dead-write-removal---consider that complete
executions of
\begin{math}
  \sem{\PW{x}{1}\SEMI\PR{x}{r}}
\end{math}
must include a read action and that complete executions of
\begin{math}
  \sem{\PW{x}{1}\SEMI\PW{x}{2}}
\end{math}
must include two write actions.  As another example, \PwTmca{} does not
validate the reverse inclusions for
\reflem{lem:if}\ref{lem:ifelse:if:if}---consider that
\begin{math}
  \sem{\IF{r}
  \THEN \PW{x}{1}
  \ELSE \PW{x}{2}
  \FI}
\end{math}
has an augmented (\reflem{lem:augment}) execution with
\begin{math}
  (r{=}0\mid\DW{x}{2})
  \xwki
  (r{\neq}0\mid\DW{x}{1}),
\end{math}
whereas
\begin{math}
  \sem{
    \IF{r} \THEN \PW{x}{1}\FI
    \SEMI
    \IF{\BANG r} \THEN \PW{x}{2}\FI
  \FI}
\end{math}
has no such execution.  We expect that these optimizations can be validated,
at the cost of complicating the semantics.  For access elimination, it is
likely sufficient to allow events with different actions to merge.  For
\reflem{lem:if}\ref{lem:ifelse:if:if}, it is likely sufficient to encoding
$\rdelay$ in the logic---the problem in the execution above is that $\rdelay$
introduces order regardless of whether preconditions are disjoint.
%\todo{Comment on \PwTpo{} and \PwTc{}?}
